%Version 3.1 December 2024
% See section 11 of the User Manual for version history
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%\documentclass[referee,sn-basic]{sn-jnl}% referee option is meant for double line spacing

%%=======================================================%%
%% to print line numbers in the margin use lineno option %%
%%=======================================================%%

%%\documentclass[lineno,pdflatex,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style

%%=========================================================================================%%
%% the documentclass is set to pdflatex as default. You can delete it if not appropriate.  %%
%%=========================================================================================%%

%%\documentclass[sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style

%%Note: the following reference styles support Namedate and Numbered referencing. By default the style follows the most common style. To switch between the options you can add or remove Numbered in the optional parenthesis.
%%The option is available for: sn-basic.bst, sn-chicago.bst%

%%\documentclass[pdflatex,sn-nature]{sn-jnl}% Style for submissions to Nature Portfolio journals
%%\documentclass[pdflatex,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
\documentclass[pdflatex,sn-mathphys-num]{sn-jnl}% Math and Physical Sciences Numbered Reference Style
%%\documentclass[pdflatex,sn-mathphys-ay]{sn-jnl}% Math and Physical Sciences Author Year Reference Style
%%\documentclass[pdflatex,sn-aps]{sn-jnl}% American Physical Society (APS) Reference Style
%%\documentclass[pdflatex,sn-vancouver-num]{sn-jnl}% Vancouver Numbered Reference Style
%%\documentclass[pdflatex,sn-vancouver-ay]{sn-jnl}% Vancouver Author Year Reference Style
%%\documentclass[pdflatex,sn-apa]{sn-jnl}% APA Reference Style
%%\documentclass[pdflatex,sn-chicago]{sn-jnl}% Chicago-based Humanities Reference Style

%%%% Standard Packages
%%<additional latex packages if required can be included here>

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
\usepackage{stmaryrd}
\usepackage{xspace}
\usepackage{amsmath}
\usepackage{mathpartir}
\usepackage{subcaption}
\usepackage{tcolorbox}
\usepackage{contour}

\contourlength{1pt}

%%%%

%%%%%=============================================================================%%%%
%%%%  Remarks: This template is provided to aid authors with the preparation
%%%%  of original research articles intended for submission to journals published
%%%%  by Springer Nature. The guidance has been prepared in partnership with
%%%%  production teams to conform to Springer Nature technical requirements.
%%%%  Editorial and presentation requirements differ among journal portfolios and
%%%%  research disciplines. You may find sections in this template are irrelevant
%%%%  to your work and are empowered to omit any such section if allowed by the
%%%%  journal you intend to submit to. The submission guidelines and policies
%%%%  of the journal take precedence. A detailed User Manual is available in the
%%%%  template package for technical guidance.
%%%%%=============================================================================%%%%

%% as per the requirement new theorem styles can be included as shown below
\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{lemma}[theorem]{Lemma}%
\newtheorem{proposition}[theorem]{Proposition}%
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%

\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}%

\raggedbottom
%%\unnumbered% uncomment this for unnumbered level heads


\include{definitions}

\begin{document}

\title{Convenient Algebraic Programming with Coercive Subtyping}

\author*[1]{\fnm{Henry} \sur{Blanchette}}
\email{blancheh@umd.edu}

\author*[2]{Aaron Stump}
\email{stumpaa@bc.edu}

\affil[1]{\orgname{University of Maryland}}
\affil[2]{\orgname{Boston College}}

% \begin{CCSXML}
%   <ccs2012>
%   <concept>
%   <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
%   <concept_desc>Software and its engineering~Functional languages</concept_desc>
%   <concept_significance>500</concept_significance>
%   </concept>
%   <concept>
%   <concept_id>10011007.10011006.10011008.10011024.10011033</concept_id>
%   <concept_desc>Software and its engineering~Recursion</concept_desc>
%   <concept_significance>500</concept_significance>
%   </concept>
%   </ccs2012>
% \end{CCSXML}

% \ccsdesc[500]{Software and its engineering~Functional languages}
% \ccsdesc[300]{Software and its engineering~Recursion}

\keywords{strong functional programming, type-based termination}

\abstract{ This paper presents a system for more convenient algebraic
  programming, where datatypes are defined as least fixed points of
  covariant signature functors, and recursive functions are defined as
  algebras. This approach requires rolling and unrolling fixed points,
  and conversion of algebras to functions that recurse over data. To
  alleviate the burden of these extra operations, in this paper we
  propose to insert them automatically, using coercive subtyping. The
  type checker generates subtyping constraints, which are then solved
  with the help of a custom logic programming engined called
  Chronolog. The subtyping axioms are given to Chronolog as rules,
  which uses them to solve the subtyping constraints. To avoid
  infinite applications of rules for unrolling signature functors,
  Chronolog provides a mechanism for suspending and resuming certain
  forms of constraints. The subtyping axioms presented to Chronolog are an
  equivalent algorithmic version of the usual declarative rules for
algebraic programming. }

\maketitle

\section{Introduction} % Aaron

% Algebraic programming as an example of total functional programming (cite Turner)
% Mendler-style algebra
% Burden of inserting roll/unroll and cata
% We will reduce this burden using coercive subtyping to infer the coercions

Strong functional programming is a form of pure functional programming
where all functions are statically required to terminate on all
inputs.  One way to achieve this is through an algebraic approach:
datatypes are least fixed points $\sMu{F}$ of signature functors $F$,
and recursive functions are least fixed points $\cata{a}$, called
catamorphisms, of algebras $a$.  The functor $F$ describes one layer
of the data, and $\sMu{F}$ then describes data consisting of any finite
number of layers.  We may unroll $\sMu{F}$ to $\sFunctorApp{F}{\sMu{F}}$, and roll it
back again.  Algebras interpret the operations declared for one layer
of the data, and catamorphisms apply algebras across all the layers.
There is no other mechanism for recursion or looping in the language,
leading  to a strong guarantee: all functions written in this
style terminate.

One burden is the need to include calls to roll and unroll functors,
and to $\cata{-}$, in code.  This adds clutter, compared to regular
functional programming.  In this paper, we address this problem by
automatically inserting those coercions wherever they are needed in
code.  The resulting code looks indistinguishable from regular
pure-functional code, and yet falls within the algebraic style, thus
assuring termination.  To insert these calls automatically, we turn to
\emph{coercive subtyping}.  The type checker generates subtyping
constraints that need to hold, in order for the term to be well-typed.
The constraints are processed by a constraint solver, generating
coercions that are then inserted at the places where the subtyping
constraints were generated.

To implement this, we make use of a custom logic-programming engine
called Chronolog.  The subtyping axioms are given to Chronolog as
rules.  Chronolog then uses standard SLD resolution to search for
proofs of the constraints, instantiating type variables during proof
search via unification.  The rules for rolling and unrolling
signatures would, if handled naively, lead to diverging proof search.
Instead, Chronolog provides a mechanism to indicate that certain
problematic constraints should be suspended.  These suspended constraints may be
resumed if they are instantiated during solving of other constraints.
If not, an outer loop of our algorithm handles them by simply equating
the lower and upper bound of the subtyping constraints.  We prove
that solving in this manner terminates.

Standard subtyping rules generally are not suitable for logic
programming, due to the presence of a transitivity rule, which
nondeterministically guesses $B$ with $A \subtype B$ and $B \subtype C$, to solve
$A \subtype C$.  We identify (Section~\ref{sec:subtyping}) an algorithmic
set of subtyping rules, without transitivity, and prove it equivalent
in Agda to the declarative rules.  This both ensures we can use logic
programming (with our algorithmic rules) to solve these subtyping
constraints, and also assures us that the algorithmic rules are indeed
correct.  To summarize the contributions:
\begin{itemize}
  \item We propose a new approach to total algebraic programming,
    where calls for rolling and unrolling signature functors
    and turning algebras into catamorphisms are inferred by coercive
    subtyping.
  \item We realize this approach with constraint-generating typing,
    and a constraint solver based on logic programming, modified
    to avoid infinite unrolling of functors. We prove termination of
    this algorithm.
  \item We propose an algorithmic subtyping system, and prove it
    equivalent to the declarative one in Agda.
\end{itemize}


\section{Syntax}

Figure~\ref{fig:syntax} presents the syntax for the language \lcata, through
which we study coercion inference for algebraic programming. We use notation
$\sepBy{e}{~\oplus}$ where $\oplus$ is a syntactic operator, to indicate a
$\oplus$-separated list of expressions $e$. Datatypes $D$ are described as least
fixed points of signature functors $F$, which we take to be polynomial functors
for simplicity. Types feature least fixed points $\sMu{D}$ of signature functors
(referenced by datatype name $D$), as well as usual function types $A \sArr B$.
Types $D \sAlg B$ are the types for algebras over $\sMu{D}$ computing a value of
type $B$, the carrier of the algebra.  Recursion type variables $\sRecVar{f}$
are introduced in the bodies of algebras, where $f$ is the name of the function
introduced by the respective algebra.

Terms include the usual constructs of simply typed $\lambda$-calculus. There are
also constructor applications $k(\sepBy{a}{,})$, and simple pattern-matching
with notation $\gamma a . \{ \sepBy{k(\sepBy{x}{,}) \mapsto b}{~|} \}$, where
the value of $a$ is matched against constructor patterns $k(\sepBy{x}{,})$.  For
totality, all constructors for $a$'s datatype must be covered. We distinguish
applications $c(a)$ of coercions $c$, which include rolling, unrolling, and
catamorphizing.  Coercions will be explained with their typing rules in
Section~\ref{sec:typechecking}.

Algebras are denoted $\omega f(x) . b$.  We here use Mendler-style
algebras~\cite{uustalu99}.  With this style, an algebra is presented with a
function $f$ from $\sRecVar{f}$ to the carrier $B$ of the algebra.
$\sRecVar{f}$ is an abstract type introduced in the algebra.  The input $x$ has
type $\sFunctorApp{D}{R}$, where $D$ is the signature functor of the algebra.
The body $b : B$ may pattern-match on $x$ and apply $f$ to subdata of type
$\sRecVar{f}$ that are revealed.  One benefit over conventional $F$-algebras is
that calling $f$ to recurse makes algebras appear very similar to conventional
recursive functional programs, albeit with more abstract typing.


\begin{figure}[htb]
  \centering

  \begin{align*}
    \begin{array}{lrcl}
      \text{(Datatype names)} &
      D &\in& \{ \sNat, \sNatList, \dots \}
      \\[0.5em]
      %
      \text{(Constructor names)} &
      k &\in& \{ \sZero, \sSuc, \sNil, \sCons, \dots \}
      \\[0.5em]
      %
      \text{(Term variables)} &
      x &\in& \textbf{TermVars}
      \\[0.5em]
      %
      \text{(Recursion type variables)} &
      % R &\in& \textbf{RecTypeVars}
      R &::=& \sRecVar{f}
      \\[0.5em]
      %
      \text{(Type variables)} &
      X, Y &\in& \textbf{TypeVars}
      \\[0.5em]
      %
      \text{(Contexts)} &
      \Gamma
      &::=&
      \cdot \mid
      \Gamma , (x : A) \mid
      \Gamma , (\sRecVar{f} : \ast)
      \\[0.5em]
      %
      \text{(Datatype signature functor)} &
      F &::=& \mu X . \sepBy{k(\sepBy{P(X)}{~\times})}{~+}
      \\[0.5em]
      %
      \text{(Constructor parameter types)} &
      P(X) &::=&
      X \mid
      \sFunctorApp{D}{P(X)} \mid
      \sMu{D}
      \\[0.5em]
      %
      \text{(Types)} &
      A, B
      & ::= &
      X \mid
      R \mid
      \sMu{D} \mid
      \sFunctorApp{D}{A} \mid
      A \sArr B \mid
      D \sAlg B
      \\[0.5em]
      %
      \text{(Terms)} &
      a, b, f
      & ::= &
      x \mid
      k(\sepBy{a}{,}) \mid
      \lambda (x : A) . b \mid
      (f ~ a) \mid
      \omega f(x) . b \mid
      \\[0.25em] &&&
      \gamma a . \{ \sepBy{k(\sepBy{x}{,}) \mapsto b}{~|} \} \mid
      \sLet{x}{A}{a}{b} \mid
      \sCoerce{c}{a}
      \\[0.5em]
      \text{(Coercions)} &
      c, c_1, c_2
      & ::= &
      \sCata{c_1}{c_2} \mid
      \sRoll{c} \mid
      \sUnroll{c} \mid
      \sSubCov{D}{c} \mid
      \\[0.25em] &&&
      \sSubData{D} \mid
      \sSubArr{c_1}{c_2} \mid
      \sSubAlg{D}{c} \mid
      \sSubRecVar{R}
    \end{array}
  \end{align*}
\caption{Syntax of \lcata}
\label{fig:syntax}
\end{figure}

Throughout the remainder of this paper, we implicitly constrain each metavariable to range exclusively over the valid syntactic forms defined by its corresponding production in Figure~\ref{fig:syntax}.

A datatype is defined by a tuple $\langle D , F \rangle$ where $D$ is
its name and $F$ is its signature functor.  This approach was proposed
and developed in early works like~\cite{hagino87,goguen+77}.  For
examples presented in this paper, we define the following set of
datatypes.
\begin{align*}
  \datatypes := \{
    & \langle \sNat , \mu X . \sZero + \sSuc(X) \rangle, \\
    & \langle \sNatList, \mu X . \sNil + \sCons(\sMu\sNat \times X) \rangle
  \}
\end{align*}

\todoHenry{note that type variables only appear in constraints, but all types inferred and that you can write in the langauge should be concrete (variable-free)}
\todoHenry{make sure that all overlines use proper separator}

\section{Motivating examples}
\label{sec:motivating-examples}

\begin{figure}[H]
  \centering

  % sub h1 expected type is μNatList -> μNatList
  % sub h1 inferred type is NatList => μNatList
  % D := NatList
  % A := μNatList
  % B := μNatList
  % C := μNatList
  % B <:: μD <--> μNatList <:: μNatList
  % A <:: C <--> μNatList <:: μNatList
  \begin{subfigure}[t]{1.0\textwidth}
  \begin{verbatim}
sub : μ Nat → Nat ⇒ μ Nat =
  λ m . ω sub'(n) . γ n .
    { Zero → m
    | Suc(n') → γ (unroll id) - (sub' n') .
        { Zero → (roll id) - (Zero)
        | Suc(m') → m' } }

subs : NatList ⇒ μ NatList → μ NatList =
  ω subs'(l1) . λ l2 . γ l1 .
    { Nil → Nil
    | Cons(h1, t1) → γ (unroll id) - (l2) .
        { Nil → (roll id) - Nil
        | Cons(h2, t2) →
          (roll id) -
            (Cons((cata id id) - sub h1 h2, subs' t1 t2)) } }
  \end{verbatim}
  \end{subfigure}
  \caption{Algebraic implementation of \texttt{sub} and \texttt{subs}, with explicit coercions. }
  \label{fig:example-subs-algebraic-with-coercions}
\end{figure}

Figure~\ref{fig:example-subs-algebraic-with-coercions} shows code with
explicit coercions, for two functions.  The \verb|sub| function
subtracts two natural numbers, and \verb|subs| proceeds through a two
lists of numbers, subtracting corresponding elements.  So subtracting
the list (using Haskell syntax) $[1,2,3]$ from $[8,9,10]$ produces
$[7,7,7]$, for example.  Let us see how these use the different kinds
of coercions.  The \verb|sub| function takes in the minuend \verb|m|
and returns an algebra, which can be used to recurse through the
subtrahend \verb|n|.  The code matches (with $\gamma$) on \verb|n|,
returning \verb|m| in the \verb|Zero| case.  In the \verb|Suc(n')|
case, we recursively subtract \verb|n'| (from \verb|m|).  Since our
algebras are Mendler-style, the type of \verb|sub'| is
\verb|R → μ Nat|, where \verb|R| is the abstract type introduced by
the algebra.  The input $n$ has type \verb|Nat R|, and so its subdatum
\verb|n'| has type \verb|R|.  So the call \verb|sub' n'| is
well-typed.  To pattern-match on the output of this call requires
unrolling \verb|μ Nat| with \verb|unroll id|.  The \verb|id| is
the identity coercion, indicating that the subtyping is only happening
at the top level of the type.  In the \verb|Zero| case of this match
on unrolled \verb|sub' n'|, the code returns \verb|Zero|.
Constructors' return types are applications of signature functors.
Here, \verb|Zero| can be given type \verb|Nat (μ Nat)|, which is then
coerced to \verb|μ Nat| by \verb|roll|.  The function \verb|subs| does
similar rolling and unrolling, but also invokes subtraction
(\verb|sub|).  Since \verb|sub| returns an algebra, the code uses
\verb|cata| to coerce that to the function which accepts the
subtrahend.

Thanks to the restrictions imposed by Mendler-style algebras, these
functions have the desirable property that they are guaranteed to
terminate on all inputs.  Nevertheless, we expect the reader will
agree that the code is rather burdened with coercions.  This
raises the question we address in this paper: can we support algebraic
programming like this with lighter syntax?  We return with our
positive answer for these examples in
Figure~\ref{fig:example-subs-algebraic-without-coercions} below.

\section{Typing}
\label{sec:typing}

\todoHenry{Write prose for introducing type system and explain how it fits into the running examples from "Motivating Examples".  Here's the idea: Here are the most interesting typing rules. The rest are in Appendix~\ref{app:typing}.
}

% Note(Henry): these are defined as macros so they can be referred to again the appendix
\newcommand{\typingRuleOmega}{
  \inferrule[Gamma]{
    \langle D, \mu X . \sepBy{k_i(\sepBy{A_{ij}}{~\times})}{~+} \rangle \in \datatypes
    \qquad
    a : \sFunctorApp{D}{C}
    \qquad
    \sepBy{\Gamma, \sepBy{x_{ij} : A_{ij}[X \mapsto C]}{,} \vdash b_i : B}{ }
  }{
    \Gamma \vdash \gamma a . \{ \sepBy{k_i(\sepBy{x_{ij}}{,}) \mapsto b_i}{~|} \} : B
  }
}
\newcommand{\typingRuleGamma}{
  \inferrule[Omega]{
    \Gamma, (\sRecVar{f} : \ast), (f : \sRecVar{f} \sArr A), (x : \sFunctorApp{D}{\sRecVar{f}}) \vdash b : B
  }{
    \Gamma \vdash \omega f(x) . b : D \sAlg B
  }
}
\newcommand{\typingRuleSubtype}{
  \inferrule[Subtype]{
    \Gamma \vdash A \subtype A' \qquad
    \Gamma \vdash a : A
  }{
    \Gamma \vdash a : A'
  }
}

\begin{figure}[htb]
  \centering

  \begin{mathpar}
    \typingRuleOmega

    \typingRuleGamma

    \typingRuleSubtype
  \end{mathpar}

  \caption{Notable typing rules}
  \label{fig:typing-rules}
\end{figure}

\section{Type checking with constraints}
\label{sec:typechecking}

\todoHenry{mention we are not doing full type inference, and are just using type variables for the sake of inferring coercions. so, lambdas and lets still have annotations}

We present a type checking procedure based on generating and solving subtyping constraints.
First, subtyping constraints are generated from a structurally recursive analysis of a term, where each constraint corresponds to a position in the term at which a coercion (which may end up being the identity) is required to justify the subtype relation.
Then, these subtyping constraints are solved by a logic-programming engine by applying the algorithmic subtyping rules in Figure~\ref{fig:algorithmic-subtyping-rules}, where each rule corresponds to a constructor of the coercion to be inserted at the constraint's associated position in the term.

\begin{figure}[htb]
  \centering

  \begin{mathpar}

    \inferrule[Var]{
      (x : A) \in \Gamma
    }{
      \Gamma \vdash x : A \gsep \varnothing
    }

    \inferrule[Con]{
      \freshMetavariables{X}
      \qquad
      \langle D, \mu Y . \dots + k(\sepBy{A'_i}{~\times}) + \dots \rangle \in \datatypes
      \qquad
      \sepBy{\Gamma \vdash a_i : A_i \gsep \constraints_{a_i}}{ }
    }{
      \Gamma \vdash k(\sepBy{a_i}{,}) : \sFunctorApp{D}{Y} \gsep \bigcup_{i} (\constraints_{a_i} \cup \{ A_i \subtypeAP{a_i} A'_i[Y \mapsto X] \})
    }

    \inferrule[Lam]{
      \Gamma, (x : A) \vdash b : B \gsep \constraints_b
    }{
      \Gamma \vdash \lambda (x : A) . b : A \sArr B \gsep \constraints_b
    }

    \inferrule[App]{
      % [Note]: We introduce a type variable for the supertype of the codomain of f so that B <:: A -> X is allowed, and then any subtyping required to make A fit is done on f rather than a
      \freshMetavariables{X} \qquad
      \Gamma \vdash f : B \gsep \constraints_f \qquad
      \Gamma \vdash a : A \gsep \constraints_a
    }{
      \Gamma \vdash f ~ a : X \gsep
      \constraints_f \cup \constraints_a \cup \{ B \subtypeAP{f} A \sArr X \}
    }

    \inferrule[Gamma]{
      % [Note]: C is the type of the thing being discriminated, which should be a subtype of D(X), where X is the type variable for the recursive parameter of D, which could be refined by C <:: D(X). And Y is the type variable for the output of the entire pattern match, which is refined by B_i <:: Y for the outputs of each branch which are the B_i
      \freshMetavariables{X, Y} \qquad
      \langle D, \mu X . \sepBy{k_i(\sepBy{A_{ij}}{~\times}) ~+} \rangle \in \datatypes
      \\\\
      \Gamma \vdash a : C \gsep \constraints_a
      \qquad
      % (\forall i)~ k_i : \sFunctorConSig{\sepBy{A_{ij}}}{,}{D}{X} \qquad
      % \\\\
      % (\forall i)~ \Gamma, \sepBy{x_{ij} : A_{ij}}{,} \vdash b_i : B_i \gsep \constraints_{b_i}
      \sepBy{\Gamma, \sepBy{x_{ij} : A_{ij}[X \mapsto C]}{,} \vdash b_i : B_i \gsep \constraints_{b_i}}{ }
    }{
      \Gamma \vdash \gamma a . \{ \sepBy{k_i(\sepBy{x_{ij}}{,}) \mapsto b_i}{~|} \} : Y
      \gsep
      \{ C \subtypeAP{a} \sFunctorApp{D}{X} \} \cup
      \constraints_a \cup
      \bigcup_i (\{ B_i \subtypeAP{b_i} Y \} \cup \constraints_{b_i})
    }

    \inferrule[Omega]{
      \Gamma, (\sRecVar{f} : \ast), (f : \sRecVar{f} \sArr A), (x : \sFunctorApp{D}{\sRecVar{f}}) \vdash b : B \gsep \constraints_b
    }{
      \Gamma \vdash \omega f(x) . b : D \sAlg B \gsep \constraints_b
    }

    \inferrule[Let]{
      \Gamma \vdash a : A \gsep \constraints_a \qquad
      \Gamma, (x : A') \vdash b : B \gsep \constraints_b
    }{
      \Gamma \vdash \sLet{x}{A'}{a}{b} : B \gsep \constraints_a \cup \{ A \subtypeAP{a} A' \} \cup \constraints_b
    }

    % [Note from Henry]: Actually this is unneeded in the subtype constraint generation rules, since coercions should only appear in solved terms.
    % % [Note from Henry]: since this is already a coercion, requires that the domain *exactly* matches the inferred type of the arg
    % \inferrule[InferCoe]{
    %   \Gamma \vdash c : \sCoerceSig{A}{B} \qquad
    %   \Gamma \vdash a : A \gsep \constraints_a
    % }{
    %   \Gamma \vdash \sCoerce{c}{a} : B \gsep \constraints_a
    % }
  \end{mathpar}

  \caption{Constraint generation rules.}
  \label{fig:constraint-generation-rules}
\end{figure}

\begin{figure}[htb]
  \centering

  \begin{mathpar}
    
    \inferrule[ReflCoe]{ }{
      \Gamma \vdash \sSubData{D} : \sCoerceSig{\sMu{D}}{\sMu{D}}
    }
        
    \inferrule[RecVarCoe]{ }{
      \Gamma \vdash \sSubRecVar{\sRecVar{f}} : \sCoerceSig{\sRecVar{f}}{\sRecVar{f}}
    }

    \inferrule[ArrCoe]{
      \Gamma \vdash c_1 : \sCoerceSig{A'}{A} \qquad
      \Gamma \vdash c_2 : \sCoerceSig{B}{B'}
    }{
      \Gamma \vdash \sSubArr{c_1}{c_2} : \sCoerceSig{A \sArr B}{A' \sArr B'}
    }

    \inferrule[CovCoe]{
      \Gamma \vdash c : \sCoerceSig{A}{A'}
    }{
      \Gamma \vdash \sSubCov{D}{c} : \sCoerceSig{\sFunctorApp{D}{A}}{\sFunctorApp{D}{A'}}
    }

    \inferrule[AlgCoe]{
      \Gamma \vdash c : \sCoerceSig{A}{A'}
    }{
      \Gamma \vdash \sSubAlg{D}{c} : \sCoerceSig{D \sAlg A}{D \sAlg A'}
    }

    \inferrule[RollCoe]{
      \Gamma \vdash c : \sCoerceSig{A}{\sMu{D}}
    }{
      \Gamma \vdash \sRoll{c} : \sCoerceSig{\sFunctorApp{D}{A}}{\sMu{D}}
    }

    \inferrule[UnrollCoe]{
      \Gamma \vdash c : \sCoerceSig{\sMu{D}}{A}
    }{
      \Gamma \vdash \sUnroll{c} : \sCoerceSig{\sFunctorApp{D}{A}}{\sMu{D}}
    }

    \inferrule[CataCoe]{
      \Gamma \vdash c_1 : \sCoerceSig{B}{D} \qquad
      \Gamma \vdash c_2 : \sCoerceSig{A}{C}
    }{
      \Gamma \vdash \sCata{c_1}{c_2} : \sCoerceSig{D \sAlg A}{B \sArr C}
    }
  \end{mathpar}

  \caption{Coercion typing rules.}
  \label{fig:coercion-typing-rules}
\end{figure}

% \begin{figure}[htb]
%   \centering
%   \begin{mathpar}
%     % Bool

%     \inferrule[DataBool]{ }{\isDatatype{\sBool}}

%     \inferrule[Bool.True]{ }{
%       \sTrue : \sFunctorConSig{}{\sBool}{A}
%     }

%     \inferrule[Bool.False]{ }{
%       \sFalse : \sFunctorConSig{}{\sBool}{A}
%     }

%     % Nat

%     \inferrule[DataNat]{ }{\isDatatype{\sNat}}

%     \inferrule[Nat.Zero]{ }{
%       \sZero : \sFunctorConSig{}{\sNat}{A}
%     }

%     \inferrule[Nat.Suc]{ }{
%       \sSuc : \sFunctorConSig{A}{\sNat}{A}
%     }

%     % NatList

%     \inferrule[DataNatList]{ }{\isDatatype{\sNatList}}

%     \inferrule[NatList.Nil]{ }{
%       \sNil : \sFunctorConSig{}{\sNatList}{A}
%     }

%     \inferrule[NatList.Cons]{ }{
%       \sCons : \sFunctorConSig{\sNat, A}{\sNatList}{A}
%     }
%   \end{mathpar}

%   \caption{Datatype constructor type inference rules.}
%   \label{fig:datatype-contructor-type-inference-rules}
% \end{figure}

Constraint generation for a term $a$ in context $\Gamma$ is performed by applying the constraint generation rules in Figure~\ref{fig:constraint-generation-rules} recursively on the structure of $a$, which upon success produces a derivation of $\Gamma \vdash a : A \mid \constraints$.
This is a usual typing judgment augmented with a constraint problem $\constraints$, where $\constraints$ is a set of constraints.
Each constraint in these sets is of the form $A \subtypeAP{a'} B$, where $a'$ is a specific sub-term of $a$, including its unique location within $a$.
The system is designed such that a solution to $\constraints$ specifies an insertion of coercions around sub-terms of $a$, where each coercion inserted at a position $a'$ corresponds to one of the original required constraints $A \subtypeAP{a'} B$.

The constraint generation phase also performs type inference, however in contrast to usual type inference and checking procedures, this procedure cannot decide a term to be mal-typed.
The constraint generation procedure will generate a constraint problem for any well-formed and well-scoped $\Gamma, a, A$.
A derivation $\Gamma \vdash a : A \mid \constraints$ merely states that $a$ has type $A$ \textit{only if} the constraint problem $\constraints$ is solvable.

For example, constraint generation for \texttt{sub : Nat → Nat ⇒ Nat} results in the following constraint problem, where $X,Y,Z,W,U$ are fresh type variables and $R$ is a fresh recursion type variable:

\todoHenry{Make sure to update this if the constraint generation rules change.}

\todoHenry{Insert term positions on constraints, just using the terms themselves}

\begin{mathpar}
  \{
    % X is domain of λ m ...
    % R is the recursive type var for omega
    % Y is the output type of sub' n
    % U is return of γ (and so is return of ω, and so is return of λ)
    % W is the recursive param of Nat for the pattern match
    % Z is the recursive param of Zero
    %
    % γ ...
    X \sArr \sNat \sAlg U \subtypeA \sMu\sNat \sArr \sNat \sAlg \sMu\sNat,
    % γ sub' n' ...
    \sFunctorApp{\sNat}{Y} \subtypeA \sFunctorApp{\sNat}{W},
    % sub' n
    R \sArr \sMu\sNat \subtypeA R \sArr \sFunctorApp{\sNat}{Y},
    % roll Zero
    \sFunctorApp{\sNat}{Z} \subtypeA U,
    % m'
    \sMu\sNat \subtypeA U
  \}
\end{mathpar}

Here is where the constraints are introduced.
\begin{itemize}
  \item The requirement $X \sArr \sNat \sAlg U \subtypeA \sMu\sNat \sArr \sNat \sAlg \sMu\sNat$ is introduced by $\gamma ~ n$.
  \item \todo{explain sources of a few other requirements -- just focus on like 3}
\end{itemize}

\todoHenry{Explain anything else necessary for understanding constraint generation}

\todoHenry{State completeness and soundness for the constraint generation rules relative to the typing rules}

\section{Subtyping} % Aaron
\label{sec:subtyping}

We begin with the declarative subtyping rules corresponding to the
coercions of Figure~\ref{fig:coercion-typing-rules}, and then
propose an algorithmic version, where all metavariables in the
premises of the rule are already present in the conclusion, and the
sizes of the constraints decrease.  This ensures that SLD resolution
will not diverge on ground constraints.  We prove that the two sets
of rules are equivalent.

\subsection{Declarative subtyping}

\newcommand{\declcata}[1]{
  \inferrule[#1]{ }{
    D \sAlg A \subtype \sMu{D} \sArr A
}}

\begin{figure}[htb]
  \centering

  \begin{mathpar}

    \inferrule[Refl]{ }
    {
      \sMu{D} \subtype \sMu{D}
    }

    \inferrule[Arr]
    {
      A' \subtype A \quad
      B \subtype B'
    }
    {
      A \sArr B \subtype A' \sArr B'
    }

    \inferrule[Cov]{
      A \subtype A'
    }{
      \sFunctorApp{D}{A} \subtype \sFunctorApp{D}{A'}
    }

    \inferrule[Alg]{
      A \subtype A'
    }{
      D \sAlg A \subtype D \sAlg A'
    }

    \inferrule[Roll]{ }{
      \sFunctorApp{D}{\sMu{D}} \subtype \sMu{D}
    }

    \inferrule[Unroll]{ }{
      \sMu{D} \subtype \sFunctorApp{D}{\sMu{D}}
    }

    \declcata{Cata}

    \inferrule[Trans]{
      A \subtype A' \quad
      A' < A''
    }{
      A \subtype A''
    }
  \end{mathpar}

  \caption{Declarative subtyping rules.}
  \label{fig:declarative-subtyping-rules}
\end{figure}

Figure~\ref{fig:declarative-subtyping-rules} shows the rules for declarative
subtyping $A \subtype B$.  We have reflexivity rule \IRule{Refl} for datatypes.  \IRule{Arr} is the usual
subtyping rule for function types, which is contravariant in the
domain and covariant in the codomain.  \IRule{Cov} expresses that the
signature functor is indeed a (covariant) functor.  \IRule{Alg}
expresses that algebra types $D \sAlg A$ are covariant in their
carriers $A$.  \IRule{Roll} and \IRule{Unroll} together express that
$\sMu{F}$ is equivalent to $\sFunctorApp{F}{\sMu{F}}$.  \IRule{Cata} expresses that
an algebra can be coerced to a function, namely the algebra's
catamorphism.  Finally, there is the \IRule{Trans} rule, completing
the definition of subtyping as a partial order.

\subsection{Algorithmic subtyping}

\newcommand{\decproccata}[1]{
  \inferrule[#1]{
    B \subtypeA \sMu{D} \quad
    A \subtypeA C
  }{
    D \sAlg A \subtypeA B \sArr C
}}

\begin{figure}[htb]
  \centering

  \begin{mathpar}

    \inferrule[Refl]{ }
    {
      \sMu{D} \subtypeA \sMu{D}
    }

    \inferrule[Arr]
    {
      A' \subtypeA A \quad
      B \subtypeA B'
    }
    {
      A \sArr B \subtypeA A' \sArr B'
    }

    \inferrule[Cov]{
      A \subtypeA A'
    }{
      \sFunctorApp{D}{A} \subtypeA \sFunctorApp{D}{A'}
    }

    \inferrule[Alg]{
      A \subtypeA A'
    }{
      D \sAlg A \subtypeA \sMu{D} \sAlg A'
    }

    \inferrule[Roll]{
      A \subtypeA \sMu{D}
    }{
      \sFunctorApp{D}{A} \subtypeA \sMu{D}
    }

    \inferrule[Unroll]{
      \sMu{D} \subtypeA A
    }{
      \sMu{D} \subtypeA \sFunctorApp{D}{A}
    }

    \decproccata{Cata}

  \end{mathpar}

  \caption{Algorithmic subtyping rules.}
  \label{fig:algorithmic-subtyping-rules}
\end{figure}

Figure~\ref{fig:algorithmic-subtyping-rules} presents the rules for
algorithmic subtyping $A \subtypeA B$.  We use similar names as for the
declarative rules, but critically, the algorithmically problematic
\IRule{Trans} rule is not present.  A consequence of this is that now
several rules that did not have premises in the declarative
formulation, here do.  For example, compare the declarative (on the
left) and algorithmic \IRule{Cata} rules:
\[
  \begin{array}{lll}
    \declcata{\ } &\ \ \ \ \ & \decproccata{\ }
  \end{array}
\]
\noindent The algorithmic rule needs to take into account subtyping
relationships between the components in the conclusion, because there
is no rule for transitivity, which could allow those relationships to
be taken into account after this rule is applied.  In effect,
transitivity needs to be built into all the other rules.  We see a
similar change from the declarative to the algorithmic versions of the
\IRule{Roll} and \IRule{Unroll} rules, for rolling and unrolling the
signature functor.

\begin{lemma}[Decrease]
  \label{lem:decrease}
  For all rules of Figure~\ref{fig:algorithmic-subtyping-rules}, the sizes
  of the constraints in the premises are less than the size of the constraint
  in the conclusion.
\end{lemma}

\begin{theorem}[Decidability]
  \label{thm:dec}
  It is decidable whether or not two types are in the algorithmic subtyping
  relation.
\end{theorem}
\begin{proof}
  Bottom-up application of the rules to ground types is guaranteed to terminate,
  by Lemma~\ref{lem:decrease}.  Therefore, proof search may be used as a decision
  procedure for algorithmic subtyping of ground types.
\end{proof}

\subsection{Equivalence of declarative and algorithmic subtyping}
\label{sec:equiv}

While Theorem~\ref{thm:dec} ensures that we can test ground subtyping
constraints using proof search, we must confirm that the algorithmic
rules are indeed a correct version of the declarative ones.  For this,
we prove the following theorems.

\begin{theorem}[Soundness]
  $A \subtypeA B$ implies $A \subtype B$.
\end{theorem}

\begin{theorem}[Completeness]
  \label{thm:completeness}
  $A \subtype B$ implies $A \subtypeA B$.
\end{theorem}

The sets of rules are quite similar, so these proofs are not onerous.  They do depend on the following
inductively proven lemmas.

\begin{lemma}[roll1 (declarative)]
  If $A \subtype \sMu{D}$, then $\sFunctorApp{D}{A} \subtype \sMu{D}$.
\end{lemma}

\begin{lemma}[roll2 (declarative)]
  If $\sMu{D} \subtype A$, then $\sMu{D} \subtype \sFunctorApp{D}{A}$.
\end{lemma}

\begin{lemma}[refl (algorithmic)]
  $A \subtypeA A$.
\end{lemma}

\begin{lemma}[trans (algorithmic)]
  If $A \subtypeA B$ and $B \subtypeA C$ then $A \subtypeA C$.
\end{lemma}

\begin{lemma}[roll (algorithmic)]
  $\sFunctorApp{D}{\sMu{D}} \subtype \sMu{D}$
\end{lemma}

\begin{lemma}[unroll (algorithmic)]
  $\sMu{D} \subtype \sFunctorApp{D}{\sMu{D}}$
\end{lemma}

\subsection{Formalization in Agda}
\label{sec:agda}

Types are represented by the Agda datatype \verb|Ty| in
Figure~\ref{fig:synagda}.  For simplicity, we have a single signature
functor \verb|D| and its least fixed-point \verb|μD| (note that this
  is a single identifier in Agda, not an application of a \verb|μ|
operator to \verb|D|).  Algebra types $D \sAlg A$ are represented just
as \verb|D⇒ A|.  Note that \verb|D⇒| is also a single symbol in Agda.  Arrow types are represented $\arr{A}{B}$ (as other
arrow notations are already in use).  The figure declares some fixity
information, for lighter notation below.

\begin{figure}
\begin{verbatim}
data Tp : Set where
  D_ : Tp → Tp
  _⟶_ : Tp → Tp → Tp
  D⇒ : Tp → Tp
  μD : Tp

infixr 9 _⟶_
infixr 10 D_
\end{verbatim}
  \caption{Agda definition of types.}
  \label{fig:synagda}
\end{figure}

Figure~\ref{fig:algagda} show the representation of the algorithmic
subtyping system; we omit a similar definition for declarative
subtyping, for space reasons.  Each set of rules is represented as an
indexed inductive type, with each rule a constructor.  For example, we
have, for both types, \verb|Refl| of type \verb|μD <: μD|, which
corresponds exactly to the \verb|Refl| rules of
Figures~\ref{fig:declarative-subtyping-rules}
and~\ref{fig:algorithmic-subtyping-rules}.  Conveniently, Agda allows
reusing constructor names across datatypes.  Dependent typing is used
for quantification.  For example, the \verb|Roll| constructor has type
\begin{verbatim}
{T : Tp} → T <:: μD → D T <:: μD
\end{verbatim}
\noindent This is a dependent function type, stating that give
\verb|T : Tp| and a proof of \verb|T <:: μD|, \verb|Roll| returns a
proof of \verb|D T <:: μD|.  Finally, the statements of soundness and
completeness are shown in Figure~\ref{fig:sndcmpl}.  The total number
of lines for all the Agda code is under 150, a very modest burden of
proof.  This is partly because types do not contain bound variables,
and so the technical overhead associated with these (for example,
with de Bruijn indices) is avoided.

\begin{figure}
\begin{verbatim}
infix 5 _<::_

data _<::_ : Tp → Tp → Set where
  Refl : μD <:: μD
  Arr : {T1 T2 T1' T2' : Tp} →
         T1' <:: T1 →
         T2 <:: T2' →
         T1 ⟶ T2 <:: T1' ⟶ T2'
  Roll : {T : Tp} → T <:: μD → D T <:: μD
  Unroll : {T : Tp} → μD <:: T → μD <:: D T
  Cov : {T1 T2 : Tp} → T1 <:: T2 → D T1 <:: D T2
  Alg : {T1 T2 : Tp} → T1 <:: T2 → D⇒ T1 <:: D⇒ T2
  Cata : {T T1 T2 : Tp} →
        T1 <:: μD →
        T <:: T2 →
        D⇒ T <:: T1 ⟶ T2
\end{verbatim}
  \caption{Agda definition of algorithmic subtyping.}
  \label{fig:algagda}
\end{figure}

\begin{figure}
\begin{verbatim}
Soundness    : {T T' : Tp} → T <:: T' → T <: T'
Completeness : {T T' : Tp} → T <: T'  → T <:: T'
\end{verbatim}
  \caption{Agda statements of soundness and completeness.}
  \label{fig:sndcmpl}
\end{figure}

\subsection{Constraint solving with Chronolog}

We take a logic-programming approach to solving problems $\constraints$ in such a way that calculates the coercions that justify the generated subtyping relations.
Our implementation uses a logic-programming engine, Chronolog, which adopts an alternative technique for constraint suspension compared to rule-level techniques in Prolog variants (e.g. Prolog II). \todo{source: previous work on delayed constraints in Prolog-style logic PLs}
Chronolog accepts a set of constraints, a logic variable substitution, and a suspension meta-predicate $S$.
Anytime a constraint that satisfies $S$ appears, that constraint is immediately suspended.
Suspended constraints are still refined by logic variable substitutions.
Chronolog can terminate in three ways: (1) failure with unsolved constraints, (2) progress with suspended constraints, (3) success with no suspended constraints.

A particular choice of $S$ is critical in this context due to the presence of the \IRule{Roll} and \IRule{Unroll} rules, which can be composed infinitely by straightforward SLD resolution, violating completeness of type checking.
Chronolog prevents such infinite regresses by supporting a global suspension meta-predicate, $S$, where constraints that satisfy $S$ are immediately suspended before being processed further.
If such a refinement results in a constraint no longer satisfying $S$, then the constraint is resumed.
% In our implementation, we use this suspension meta-predicate:
We use a meta-predicate for $S$ that holds of the following forms of constraints.
\begin{mathpar}
  X \subtypeA \sMu{D}, X \subtypeA \sFunctorApp{D}{A}, X \subtypeA R, \sMu{D} \subtypeA X, \sFunctorApp{D}{A} \subtypeA X, R \subtypeA X, X \subtypeA Y
\end{mathpar}
Suspending constraint of these forms prevents Chronolog from applying rules in a cycle, which follows from a case analysis of the all possible rule applications to constraints not of suspended forms.
In particular, the only forms of constraints that have a type variable as one side of the relation that are not suspended are those were the other side of the relation is either $D \sAlg A$ or $A \sArr B$.
This allows Chronolog, for example, to attempt to solve constraints of the form $X \subtypeA A \sArr B$ with both $\IRule{Arr}$ and $\IRule{Cata}$, or to attempt to solve constraints of the form $A \sAlg B \subtypeA X$ with both $\IRule{Alg}$ and $\IRule{Cata}$.

\todoHenry{mention that \Func{Solve} is nondeterministic}

\begin{algorithm}[htb]
  \caption{Constraint solving with Chronolog.}
  \label{alg:solver}
  \begin{algorithmic}[1]
    \Function{Solve}{$\constraints$}
    \State $\sigma \gets \emptyset$ % \Comment{Initialize as the empty logic variable substitution}
    \While{$\constraints \neq \emptyset$}
    \State $\mathit{result} \gets \Func{Chronolog}(S, \constraints, \sigma$)
    % \Statex
    % Pattern-matching block
    \State \Match $\mathit{result}$ \With
    \State \quad \Case $\langle \text{``failure''} \rangle \Rightarrow$ \Return $\bot$
    % \Statex
    \State \quad \Case $\langle \text{``success''}, \sigma' \rangle \Rightarrow$
    \Return $\sigma'$
    % \Statex
    \State \quad \Case $\langle \text{``progress''}, \sigma', \constraints' \rangle \Rightarrow$
    \State \qquad $c \gets \Func{Sample}(\constraints')$
    \State \qquad $\sigma \gets \Func{Refine}(c) \circ \sigma'$
    \State \qquad $\constraints \gets \sigma(\constraints')$
    \State \EndMatch
    \EndWhile
    \EndFunction
  \end{algorithmic}
\end{algorithm}

Constraints solving is accomplished by using Chronolog inside a loop that processes any suspended constraints left over after each iteration, as shown in Algorithm~\ref{alg:solver}.
Each iteration, a single suspended constraint is chosen to be processed by \Func{Refine}, resulting in a substitution that refines that suspended constraint (and may incidentally refine other suspended constraints).
The choice of which suspended constraint to process may affect the order in which Chronolog explores solutions but all choices lead to equivalent solutions up to permutations of \IRule{Roll} and \IRule{Unroll} (if there is a solution).
\todoHenry{justify this statement}

\begin{mathpar}
  \begin{array}{rlcrll}
    \Func{Refine}(\sMu{D} &\subtypeA X) &=& \Func{Refine}(X &\subtypeA \sMu{D}) &= \{ X \mapsto \sMu{D} \} \\
    \Func{Refine}(\sFunctorApp{D}{A} &\subtypeA X) &=& \Func{Refine}(X &\subtypeA \sFunctorApp{D}{A}) &= \{ X \mapsto \sFunctorApp{D}{Y} \} ~\text{where}~ Y ~\text{fresh} \\
    \Func{Refine}(R &\subtypeA X) &=& \Func{Refine}(X &\subtypeA R) &= \{ X \mapsto R \} \\
    &&& \Func{Refine}(X &\subtypeA Y) &= \{ X \mapsto Y \}
  \end{array}
\end{mathpar}

\begin{lemma}[Termination of \Func{Solve}]
  \label{lem:solver-termination}
  For any constraint problem $\constraints$, $\Func{Solve}(\constraints)$ (Algorithm~\ref{alg:solver}) terminates.
\end{lemma}
\begin{proof}
  We show termination by demonstrating a strictly decreasing metric over each iteration of the loop in \Func{Solve}.
  Each iteration begins with an invocation of \Func{Chronolog}.
  If \Func{Chronolog} results in $\langle \text{``failure''} \rangle$ or $\langle \text{``success''} \rangle$, then we are done.
  If \Func{Chronolog} results in $\langle \text{``progress''}, \sigma', \constraints' \rangle$, then by Lemma~\ref{lem:decrease} the leftover $\constraints'$ must be in total smaller than the original $\constraints$ from the beginning of this iteration.

  The substitution resulting from $\Func{Refine}(\textit{constraint})$ preserves the size of constraints in all but one case.
  When $\Func{Refine}(\textit{constraint})$ increases the size of constraints in one case, where it results in $\{ X \mapsto \sFunctorApp{D}{Y} \}$, \Func{Chronolog} will strictly decrease the size of these constraints again immediately, as follows from a case analysis on all the possible results of applying $\Func{Refine}(\textit{constraint})$ to suspended constraints.
\end{proof}

\begin{lemma}[Completeness of \Func{Solve}]
  \label{lem:solver-completeness}
  If $\sigma$ is a solution to a constraint problem $\constraints$, then $\exists \sigma' \in \eqClass{\sigma}{\sim}$ such that $\sigma' \in \Func{Solve}(\constraints)$ (Algorithm~\ref{alg:solver}).
\end{lemma}
\begin{proof}
  \Func{Solve} will finds all solutions (up to permutations of \IRule{Roll} and \IRule{Unroll}) to $\constraints$ because \Func{Chronolog} attempts every valid application of the algorithmic subtyping rules which are complete by Theorem~\ref{thm:completeness}, \Func{Refine} preserves validity of solution $\sigma$, and \Func{Solve} terminates by theorem \ref{lem:solver-termination}.
\end{proof}

While there may be multiple solutions to $\constraints$, solutions to such constraint problems generated by the procedure described in Section~\ref{sec:typechecking} are equivalent up permutations of \IRule{Roll} and \IRule{Unroll} \todoHenry{justify this claim}.
Furthermore, solutions related by this equivalence, when applied as coercion insertions to terms, result in semantically equivalent terms where \texttt{roll} and \texttt{unroll} are semantically interpreted as identity functions \todoHenry{justify this claim}.

\begin{algorithm}[htb]
  \caption{Type checking for \lcata.}
  \label{alg:type-checker}
  \begin{algorithmic}[1]
    \Function{TypeCheck}{$\Gamma, a, A$}
    \State $\constraints \gets \Func{Generate}(\Gamma, a, A)$
    \State \Match $\Func{Solve}(\constraints)$ \With
    \State \quad \Case $[~] \Rightarrow \bot$
    \State \quad \Case $\_ \Rightarrow \top$
    \State \EndMatch
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\begin{theorem}[Completeness of \Func{TypeCheck}]
  \label{thm:type-checker-completeness}
  If $\Gamma \vdash a : A$ according to the declarative typing rules for \lcata displayed in Figure~\ref{fig:typing-rules}, then $\Func{TypeCheck}(\Gamma, a, A)$ (Algorithm~\ref{alg:type-checker}) accepts.
\end{theorem}
\begin{proof}
  \todoHenry{Essentially, by Theorem~\ref{thm:completeness} and Lemma~\ref{lem:solver-completeness}.}
\end{proof}

\section{Returning to the Examples} % Henry

Recall the examples from Section~\ref{sec:motivating-examples} where \texttt{sub} and \texttt{subs} were implemented algebraically.
Explicit coercions using \texttt{unroll}, \texttt{roll}, and \texttt{cata} were required to operate with the representation of datatypes as fixed points of signature functors and the functional interpretation of algebras via a catamorphism.
Now, with the machinery of coercive subtyping in place, we can write those examples without any coercions, as demonstrated in Figured~\ref{fig:example-subs-algebraic-without-coercions}.

\begin{figure}[H]
  \centering

  \begin{verbatim}
sub : μ Nat → Nat ⇒ μ Nat
sub = λ (m : μ Nat) . ω sub'(n) . γ n .
  { Zero → m
  | Suc(n') → γ sub' n' .
    { Zero → Zero
    | Suc(m') → m' } }

subs : NatList ⇒ μ NatList → μ NatList
subs = ω subs'(l1) . λ (l2 : μ NatList) . γ l1 .
  { Nil → Nil
  | Cons(h1, t1) → γ l2 .
    { Nil → Nil
    | Cons(h2, t2) →
        Cons(sub h1 h2, subs' t1 t2) } }
  \end{verbatim}
  \caption{Code for \texttt{sub} and \texttt{subs} with coercions all implicit.}
  \label{fig:example-subs-algebraic-without-coercions}
\end{figure}

The type checking procedure described in previous sections collects and solves the implied subtyping constraints, which determine an insertion of coercions into these terms which matches the manual usage of explicit coercions in the original examples.

\section{Related Work} % Aaron

Turner introduced the term \emph{strong functional programming},
presented several arguments in favor of the approach, and proposed
using structural termination as an \emph{elementary} way to realize
such a language~\cite{turner95}.  Mendler proposed a recursor using an
abstract type to ensure that recursive calls are permitted only on
subdata for an inductive type~\cite{mendler91}.  The algebraic
approach has been applied for modular implementation of
interpreters~\cite{swierstra08}.  Charity is a strong functional
programming language based on categorical abstractions for initial
algebras and final coalgebras~\citep{charity92}. Tofu uses the
Calculus of Constructions as a language for defining terminating
functions, again based on categorical ideas~\citep{tofu10}.  Charity
and Tofu both support coinductive datatypes, which we did not study
here.

\section{Conclusion and future work} % Aaron

Total algebraic programming can be made more feasible by inferring the
basic coercions necessary for the method: rolling and unrolling
signature functors, and turning an algebra into its catamorphism.
Type inference generates subtyping constraints, whose solutions
correspond to the needed coercions.  A logic-programming engine
processes constraints with respect to an algorithmic version of the
subtyping rules.  This algorithmic subtyping relation is proved
equivalent to the declarative one, in Agda.

Future work includes richer forms of algebraic programming.  One
powerful generalization of structural recursion is to recursive
coalgebras~\cite{capretta+06}.  This encompasses divide-and-conquer
algorithms, which are beyond the scope of structural recursion.  A
different approach to divide-and-conquer recursion has also been
considered by~\citet{abreu+23}.  There, algebras call functions to
pass between various abstract types.  These would be excellent
candidates for inference using coercive subtyping, as we proposed in
this paper.  It would also be interesting to infer coercions for
coalgebraic programming with coinductive types.

\appendix

\section{Typing rules}
\label{app:typing}

Section~\ref{sec:typing} presented a sampling from the full set of typing rules for \lcata below.

\begin{mathpar}

  \inferrule[Var]{
    (x : A) \in \Gamma
  }{
    \Gamma \vdash x : A
  }
  
  \inferrule[Con]{
    \langle D , \mu X . \cdots + k(\sepBy{A_i}{~\times}) + \cdots \rangle \in \datatypes \qquad
    \sepBy{\Gamma \vdash a_i : A_i}{ }
  }{
    \Gamma \vdash k(\sepBy{a_i}{,}) : \sFunctorApp{D}{X}
  }
  
  \inferrule[Lam]{
    \Gamma , (x : A) \vdash b : B
  }{
    \Gamma \vdash \lambda (x : A) . b : A \sArr B
  }
  
  \inferrule[App]{
    \Gamma \vdash f : A \sArr B \qquad 
    \Gamma \vdash a : A
  }{
    \Gamma \vdash f ~ a : B
  }

  \typingRuleGamma
  
  \typingRuleOmega
  
  \inferrule[Let]{
    \Gamma, (x : A) \vdash b : B \qquad 
    \Gamma \vdash a : A
  }{
    \Gamma \vdash \sLet{x}{A}{a}{b} : B
  }

  \typingRuleSubtype

\end{mathpar}

% Richer patterns of recursion, as in POPL 2023 paper

%\bibliographystyle{plain}
\bibliography{paper.bib}

\end{document}
